---
title: "221203_Girst attempt_NO3_1000 lakes"
output: html_document
date: "2022-12-03"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r FUNCTIONS}
#To make plots individually to individual optimization
#Function for å definere antall siffer i akse
fmt_dcimals <- function(decimals=1){
  function(x) format(x,nsmall = decimals,scientific = FALSE)
}

samvif <- function(mod){
  # mod is an mgcv object
  # this function calculates the variance inflation factors for GAM as no one else has written code to do it properly
  # this is used to summarise how well the GAM performed
  mod.sum <- summary(mod)
  s2 <- mod$sig2 # estimate of standard deviation of residuals
  X <- mod$model # data used to fit the model
  n <- nrow(X) # how many observations were used in fitting?
  v <- -1 # omit the intercept term, it can't inflate variance
  varbeta <- mod.sum$p.table[v,2]^2 # variance in estimates
  varXj <- apply(X=X[,row.names(mod.sum$p.table)[v]],MARGIN=2, var) # variance of all the explanatory variables
  VIF <- varbeta/(s2/(n-1)*1/varXj) # the variance inflation factor, obtained by rearranging
  # var(beta_j) = s^2/(n-1) * 1/var(X_j) * VIF_j
  
  VIF.df <- data.frame(variable=names(VIF),
                       vif=VIF, 
                       row.names=NULL)
  return(VIF.df)
}
```

```{r Load Packages}
Packages <- c("mgcv","ggplot2", "dplyr", "viridis", "bestNormalize", "GGally", "bestNormalize", "rnaturalearth")
lapply(Packages, library, character.only = TRUE)

#virdis for colour https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
```

```{r Load data}
dfx <- read.table("WChem_1000 lakes data 1995 and 2019.txt", sep = "\t", header = TRUE)
dfy <- read.table("Runoff_1000lakes_catchment characteristics.txt", sep = "\t", header = TRUE)
df <- merge(dfx, dfy, by="Station.Code")
```

## Define natural catchments
From James: Excluding 108 catchments with combined (urban + agricultural) land cover of >= 1 %
In our data there are agriculture and developed, but no urban. Using the same criteria (1%) 153 catchments are removed
Q: should we also exclude by the other criteria used by James? catchment size etc.
```{r Extract only those considered "natural}
df$anthro = df$agriculture + df$developed
df1 <- subset(df, anthro < 1)
```

## Variables of main interest
- longitude
- latitude
- altitude
- forest
- forest/forest+natural_not_forest
- peat
- lake to catchment area 
- water flow
```{r Create new variables}
df1$forest_to_not <- df1$forest/(df1$forest+df1$natural_not_forest)
```

## Explore the data
Looking for
- outliers
- collinearity between variables
https://www.r-graph-gallery.com/89-box-and-scatter-plot-with-ggplot2.html
```{r Explore the data visually}
#Looking for outliers by different groups of variables
ggplot(df1, aes(x=Region.4, y=X2019_NO3.N)) +
    geom_boxplot() +
    geom_jitter(color="black", size=0.4, alpha=0.9)+
    scale_y_continuous(limits = c(), labels = fmt_dcimals(1))+
    labs(title = "", y ="NO3-N (ug/L)")

ggplot(df1, aes(x=Region.4, y=runoff_mm.yr)) +
    geom_boxplot() +
    geom_jitter(color="black", size=0.4, alpha=0.9)+
    scale_y_continuous(limits = c(), labels = fmt_dcimals(1))+
    labs(title = "", y ="")
```

Collinearity
- checking for collinearity
- highly corrleates not in same model

For NO3-N:
  Lengde-breddemeter
  Høyde
  (skog/skog+natural_not_forest)
  Lake-catchment area
  Vannføring

```{r collinearity}
library(GGally)
df2 <- subset(df1, select=c(latitude.x,longitude.x, altitude,
                            forest_to_not,
                            lake_catchm_., 
                            Res_time_years, runoff_mm.yr))
ggpairs(df2,title="NO3-N")
# Can we ignore those below 0.5?
#ALtidtude correlates strongly with forest-parameter, and altitude shoule be excluded from model
#Latitude and longiute strongly correlated
```

Possible collinearity between
- lat and long
- forest/forest_to_not.. and altitude (0.548)
```{r Mapping the observations}
world <- ne_countries(scale = "medium", returnclass = "sf")
ggplot(data = world, zoom=4) +
  geom_sf()+
  coord_sf(xlim = c(2, 30), ylim = c(58, 71))+
  geom_point(aes(x = longitude.x, y = latitude.x, colour=X2019_NO3.N), 
             data=df1, size =2)
#  labs(title = "", x = "", y = "")+
#  theme(axis.text.x = element_text(angle=0, vjust=1, size = 12),
#        axis.text.y = element_text(size= 12),
#        legend.title = element_text(size = 18),
#        axis.title.x = element_text(size = 13, margin=margin(13,0,0,0)),
#        axis.title.y = element_text(size = 13, margin=margin(0,13,0,0)),
#        legend.text=element_text(size=16),
#        plot.title = element_text(size = 24),
#        legend.key = element_rect(fill = "black", color = NA),
#        panel.background=element_rect(fill = "white", colour = "white"),
#        panel.border = element_rect(colour = "grey", fill=NA))
```

Looking at the distribution of the data
- should the data be normalised?
- only predictor
- Valiente et al 2022 used transformation on all variables
- what about scaling? e.g. divide latitude by 10
- or simply find the right family in the gam model
```{r Histogram of variables}
hist(df1$X2019_NO3.N)
hist(log(df1$X2019_NO3.N))
hist(orderNorm(df1$X2019_NO3.N))
hist(df1$forest_to_not)
hist(df1$Res_time_years)
hist(df1$lake_catchm_.)
hist(df1$runoff_mm.yr)
```

```{r Checking data distribution}
bestNormalize(df1$X2019_NO3.N)
bestNormalize(df1$peat)
```

INFO ON PARAMETER SELECTION GAM
- interaction with same scale, use s()
- only interaction at different scales, use te()
- interactions and signe parameters on different scales, use ti()
- recommend using method =REML or ML
- not always correct to set the degrees of freedom of smooth (k). Could check to see its not too low. Make suffiiently high. Could be as high as 100
- log link will restrict negative values
- family Gamma since left skewed?
- family scat() less sensitive to extremes (?)
- select =TRUE should be applied
- spline (bs) should be ds or tp
https://noamross.github.io/gams-in-r-course/chapter3
https://osf.io/wgc4f/wiki/mgcv:%20model%20selection/
https://noamross.github.io/gams-in-r-course/chapter3

PROCEDURE
- first make simple model including significant terms
- adjust k
- include interactions to see if improvement

MODEL CHECKING
- in summary.gam the p values are of the null hypothesis of a zero effect of the indicated spline. There values relate to the F statistic in the table produced by summary.gam,
- in gam.check the p values are for the test of the null hypothesis that the basis dimension used is of sufficient size. I.e. these p values relate to the value labelled k-index in the table produced by gam.check
- compare models using AIC. The model which has lower AIC value than the other is better than the other in the sense that it is less complex but still a good fit for the data.


```{r Decide on the major settings}
#using all parameters with no transformation. Using ts since faster than ds
NO3.1 <- gam(X2019_NO3.N ~ s(latitude.x, longitude.x)+
                  s(forest_to_not)+s(Res_time_years)+
                  s(lake_catchm_.)+s(runoff_mm.yr),
                family=scat(),
                type="response",
                select=TRUE,
                data=df1)
summary(NO3.1)
gam.check(NO3.1)

#From first model, lake_catchment not significant. R-adj = 0.414
NO3.2 <- gam(log(X2019_NO3.N) ~ s(latitude.x, longitude.x)+
                  s(forest_to_not)+s(Res_time_years)+
                  s(lake_catchm_.)+s(runoff_mm.yr),
                family=scat(),
                type="response",
                select=TRUE,
                data=df1)
summary(NO3.2)
gam.check(NO3.2)
#When log-transforming the response, lake_catchm_. is signfiicant at p<0.05. R2-adj = 0.565. Decides to keep log-transformation and not to #NO3.3 <- gam(log(X2019_NO3.N) ~ s(latitude.x, longitude.x)+
                  s(forest_to_not)+s(Res_time_years)+
                  s(runoff_mm.yr),
                family=scat(),
                type="response",
                select=TRUE,
                data=df1)

summary(NO3.3)
gam.check(NO3.3)

AIC(NO3.1, NO3.2, NO3.3)
#Very similar fit from removing lake_catchm, perhaps should be removed? Keep for now given slightly better R-adj and AIC.


#Happy with the 3rd model. From gam.check() a few of the variables are significant while we want them not to be. Can help to increase k. 
NO3.4 <- gam(log(X2019_DOC.C) ~ s(latitude.x, longitude.x, k=100)+
                  s(forest_to_not, k=15)+s(Res_time_years)+
                  s(lake_catchm_.)+
                  s(runoff_mm.yr),
                family=scat(),
                type="response",
                select=TRUE,
                data=df1)
gam.check(NO3.4)
summary(NO3.4)
#Now it looks good with regards to gam.check() but now lake_catchment is not significant at p<0.05!
NO3.5 <- gam(log(X2019_DOC.C) ~ s(latitude.x, longitude.x, k=100)+
                  s(forest_to_not, k=15)+s(Res_time_years)+
                  s(runoff_mm.yr),
                family=scat(),
                type="response",
                select=TRUE,
                data=df1)
gam.check(NO3.5)
summary(NO3.5)

plot(NO3.5, page=2, resid = TRUE)
#Now everything looks good from summary and gam.check(). R-adj 0.854

#would gamma be restricted to only positive?
# hva betyr det for oss å bruke bs="ds"?
# can not have zero values
#which identifty link?
#Parameter + parameter-interaction selection
# k adjustment

```


```{r See effect frm changing certain settings: log}
#The following was a poorer option to the log transformation of y

#link = log. If to be used we need to remove log from y
#NO3.5x <- gam((X2019_NO3.N) ~ s(latitude.x, longitude.x, k=100)+
#                  s(forest_to_not, k=15)+s(Res_time_years)+
#                  s(runoff_mm.yr),
#                family=scat(link="log"),
#                type="response",
#                select=TRUE,
#                data=df1)
#gam.check(NO3.5x)
#summary(NO3.5x)
#AIC(TOC4, TOC5)
```

```{r See effect frm changing certain settings: ds versus tp}
#The following did not improve
#NO3.6 <- gam((X2019_NO3.N) ~ s(latitude.x, longitude.x, k=100)+
#                  s(forest_to_not, k=15)+s(Res_time_years)+
#                  s(runoff_mm.yr),
#                family=scat(link="log"),
#                type="response",
#                select=TRUE,
#                data=df1)
#gam.check(NO3.6)
#summary(NO3.6)
#AIC(NO3.5, NO3.6)
#Takes so very long time to run with "ds"  in each. Did not wait for it to finish. 
```

Make plot to evaluate the wiggliness. 
```{r}
library("gratia")
draw(NO3.5, residuals = TRUE)
```



To decide on interactions and variables
```{r}
TOC7 <- gam(log(X2019_DOC.C) ~ s(latitude.x, longitude.x, k=100)+
                  s(forest_to_not)+s(peat)+s(Res_time_years)+
                  s(runoff_mm.yr)+
              ti(forest_to_not, peat),
                family=scat(),
                type="response",
                select=TRUE,
                data=df1)

#TOC8 <- gam(log(X2019_DOC.C) ~ s(latitude.x, longitude.x, k=100)+
#                  s(forest_to_not)+s(peat)+s(Res_time_years)+
#                  s(runoff_mm.yr)+
#                ti(forest_to_not, peat)+
#                ti(Res_time_years, runoff_mm.yr),
#                family=scat(),
#                type="response",
#                select=TRUE,
#                data=df1)

#interaction between res time and runoff is not significant
gam.check(TOC8)
summary(TOC8)

AIC(TOC7, TOC5)
#Model7, with interaction forest and peat, has much lower AIC than without
plot(TOC7, page=2)
```

Plotting model in map with ggplot - have not managed successfully
could be of use(?): https://mfasiolo.github.io/mgcViz/articles/mgcviz.html
```{r Plotting the model in map with ggplot}
library(tidymv)
library(mgcViz)
#model data needs to be extractes
TOC_plot <- predict_gam(TOC_all3)
b <- getViz(TOC_all3)

plot(sm(b, 1)) + l_fitRaster() + l_fitContour() + l_points()

print(plot(b, allTerms = T), pages = 1)
plot(b)
coef(TOC_all)
# https://mfasiolo.github.io/mgcViz/articles/mgcviz.html
library(rgl)
plotRGL(sm(b, 1), fix = c("z" = 0), residuals = TRUE)
```

